\replaced[id=siyu]{Separating random and deterministic sources of computational noises in explore-exploit decisions}{The nature of decision noise in random exploration}

### 2023-12-21 manuscript/2022 unused - Plos Comp Bio/manuscript_DeterministicRandomNoise.tex
u

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{We built a novel computational model which could detect and separate previously unobserved deterministic processes from random noise, without having to make a priori assumptions about the nature of the deterministic process. } In particular, we find evidence that around 15$\%$ of the variability in random exploration can be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Our work takes us one step closer to understanding of the nature of decision noise in random exploration.}

>>>>

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
		%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
		A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{We built a novel computational model which could detect and separate previously unobserved deterministic processes from random noise, without having to make a priori assumptions about the nature of the deterministic process. } In particular, we find evidence that around 15$\%$ of the variability in random exploration can be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Moreover, our results suggest that both deterministic noise and random noise change proportionally to each other as the planning horizon changes. This sheds light on a common noise gating mechanism that is at play in random exploration. Our work takes us one step closer to understanding of the nature of decision noise in explore-exploit decisions.}
